---
title: "Stewart Potentials: a Use Case"
author: "Timothée Giraud"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stewart Potentials: a Use Case}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The Stewart potentials method relies on spatial interaction modeling, it aims to compute indicators based on stock values processed in a way that take into account the units neighborhood.
These indicators have two main benefits: 

1. they make the maps more understandable in simplifying the representation of spatial patterns;  
2. they enrich the information displayed in addressing accessibility issues together with regional structural characteristics.  

At the European scale, this functional semantic simplification may help to show a smoothed context-aware picture of the localized socio-economic activities.

In this vignette, we will show a use case of these potentials on the regional GDP per capita at the european scale and we will build three maps to show the benefits of this method:

* a regional map of the GDP per capita;  
* a regional map of the potential GDP per capita;
* a smoothed map of the GDP per capita.  

Note that this example is based on data and mapping functions proposed in the **[cartography]( https://cran.r-project.org/package=cartography) package**.  


## Regional Map of the GDP per Capita
```{r regionalmap, fig.width=7, fig.height=6}
library(cartography)
library(SpatialPosition)
data(nuts2006)

# Create the GDP per capita variable
nuts3.df$gdpcap <- nuts3.df$gdppps2008 * 1000000 / nuts3.df$pop2008
# Discretization of the variable
bv <- quantile(nuts3.df$gdpcap, seq(from = 0, to = 1, length.out = 9))

# Creation of the map
opar <- par(mar = c(0,0,1,0))

# Build a color palette
pal <- carto.pal(pal1 = "wine.pal", n1 = 8)

# Set the basemap
plot(nuts0.spdf, add = F, border = NA, bg = "#cdd2d4")
plot(world.spdf, col = "#f5f5f3ff", border = "#a9b3b4ff", add = TRUE)

# Plot the regional GDP per capita
choroLayer(spdf = nuts3.spdf, df = nuts3.df, 
           var = "gdpcap", 
           legend.pos = "topright",
           breaks = bv, col = pal, 
           border = NA, 
           legend.title.txt = "GDP per capita",
           legend.values.rnd = -2, 
           add = TRUE)
plot(nuts0.spdf, add = TRUE, lwd = 0.5, border = "grey30")
plot(world.spdf, col = NA, border = "#7DA9B8", add = TRUE)

# Set a layout
layoutLayer(title = "Wealth Inequality in Europe", 
            sources = "Basemap: UMS RIATE, 2015 - Data: Eurostat, 2008", 
            author = "T. Giraud, 2015")
par(opar)
```

## Regional Map of the Potential GDP per Capita
We will compute potentials of population and GDP in each units. 
Value computed in each unit take into account the values of its neighbouring units. 

```{r regionalmappot, fig.width=7, fig.height=6 }
# Create a distance matrix between units
mat <- CreateDistMatrix(knownpts = nuts3.spdf, 
                        unknownpts = nuts3.spdf)

# Merge the data frame and the SpatialPolygonsDataFrame
nuts3.spdf@data <- nuts3.df[match(nuts3.spdf$id, nuts3.df$id),]

# Compute the potentials of population per units
# function = exponential, beta = 2, span = 75 km
poppot <- stewart(knownpts = nuts3.spdf, 
                  unknownpts = nuts3.spdf, 
                  matdist = mat,
                  varname = "pop2008", 
                  typefct = "exponential", 
                  beta = 2, 
                  span = 75000)

# Compute the potentials of GDP per units
# function = exponential, beta = 2, span = 75 km
gdppot <- stewart(knownpts = nuts3.spdf, 
                  unknownpts = nuts3.spdf, 
                  matdist = mat,
                  varname = "gdppps2008", 
                  typefct = "exponential", 
                  beta = 2,
                  span = 75000)

# Create a data frame of potential GDP per capita
pot <- data.frame(id = nuts3.df$id, 
                  gdpcap = gdppot$OUTPUT * 1000000 / poppot$OUTPUT, 
                  stringsAsFactors = FALSE)

# Creation of the map
par <- par(mar = c(0,0,1,0))

# Set the basemap
plot(nuts0.spdf, add = F, border = NA, bg = "#cdd2d4")
plot(world.spdf, col = "#f5f5f3ff", border = "#a9b3b4ff", add = TRUE)

# Plot the regional potential of GDP per capita
choroLayer(spdf = nuts3.spdf, df = pot, 
           var = "gdpcap", 
           legend.pos = "topright",
           breaks = bv, col = pal, 
           border = NA,
           legend.title.txt = "Potential\nGDP per capita",
           legend.values.rnd = -2, add = TRUE)
plot(nuts0.spdf, add=T, lwd = 0.5, border = "grey30")
plot(world.spdf, col = NA, border = "#7DA9B8", add=T)

# Set a text to explicit the function parameters
text(x = 6271272, y = 3743765, 
     labels = "Distance function:\n- type = exponential\n- beta = 2\n- span = 75 km", 
     cex = 0.8, adj = 0, font = 3)

# Set a layout
layoutLayer(title = "Wealth Inequality in Europe", 
            sources = "Basemap: UMS RIATE, 2015 - Data: Eurostat, 2008", 
            author = "T. Giraud, 2015")
par(opar)
```
This map gives a smoothed picture of the spatial patterns of wealth in Europe while keeping the original spatial units as interpretive framework. Hence, the map reader can still rely on a known territorial division to develop its analyses.


## Smoothed Map of the GDP per Capita
In this case, the potential GDP per capita is computed on a regular grid.
```{r smoothedmappot, fig.width=7, fig.height=6}
# Compute the potentials of population on a regular grid (50km span)
# function = exponential, beta = 2, span = 75 km
poppot <- stewart(knownpts = nuts3.spdf, 
                  varname = "pop2008", 
                  typefct = "exponential", 
                  span = 75000, 
                  beta = 2, 
                  resolution = 50000, 
                  mask = nuts0.spdf)

# Compute the potentials of GDP on a regular grid (50km span)
# function = exponential, beta = 2, span = 75 km
gdppot <- stewart(knownpts = nuts3.spdf, 
                  varname = "gdppps2008", 
                  typefct = "exponential", 
                  span = 75000, 
                  beta = 2, 
                  resolution = 50000, 
                  mask = nuts0.spdf)

# From the regularly spaced SpatialPointsDataFrame to a raster
popras <- rasterStewart(poppot)
gdpras <- rasterStewart(gdppot)

# GDP per capita 
ras <- gdpras * 1000000 / popras

# Create a SpatialPolygonsDataFrame from the raster
pot.spdf <- contourStewart(x = ras, 
                           breaks = bv, 
                           mask = nuts0.spdf, 
                           type = "poly")

# Creation of the map
par <- par(mar = c(0,0,1,0))

# Set the basemap
plot(nuts0.spdf, add = F, border = NA, bg = "#cdd2d4")
plot(world.spdf, col = "#f5f5f3ff", border = "#a9b3b4ff", add = TRUE)

# Plot the potential GDP per Capita
choroLayer(spdf = pot.spdf, df = pot.spdf@data, var = "mean", 
           legend.pos = "topright",
           breaks = bv, col = pal, add=T, 
           border = "grey90", lwd = 0.2,
           legend.title.txt = "Potential\nGDP per capita",
           legend.values.rnd = -2)
plot(nuts0.spdf, add=T, lwd = 0.5, border = "grey30")
plot(world.spdf, col = NA, border = "#7DA9B8", add=T)

# Set a text to explicit the function parameters
text(x = 6271272, y = 3743765, 
     labels = "Distance function:\n- type = exponential\n- beta = 2\n- span = 75 km", 
     cex = 0.8, adj = 0, font = 3)

# Set a layout
layoutLayer(title = "Wealth Inequality in Europe", 
            sources = "Basemap: UMS RIATE, 2015 - Data: Eurostat, 2008", 
            author = "T. Giraud, 2015")
par(opar)
```
Unlike the previous maps, this one omits the initial territorial division to give a smoothed and continuous picture of the spatial patterns of wealth in Europe. It eases the interpretation toward a vision of the space free from territorial divisions and addresses the well-known issues of the MAUP. 




